\documentclass[twocolumn]{aastex631}
\usepackage{amsmath}
\usepackage{graphicx}

\begin{document}

\title{Learning the Orbital-Uncertainty Propagation Operator: A Neural
Surrogate for Monte-Carlo Impact Monitoring of Near-Earth Asteroids}

\author{...}
\affiliation{...}

\begin{abstract}
Impact monitoring of near-Earth asteroids (NEAs) relies on propagating the
orbit-determination uncertainty into the future, traditionally by integrating
thousands of virtual asteroids (clones) sampled from the orbital covariance
with high-fidelity $N$-body dynamics (e.g.\ JPL Sentry). We present a neural
surrogate that learns this uncertainty-propagation operator directly: given the
cometary orbital elements and the $6\times6$ covariance matrix of an NEA, a
mixture-density network (MDN) deep ensemble predicts the full probability
distribution of the minimum Earth-approach distance in ten consecutive
decadal windows spanning 100~yr. Ground truth is generated with REBOUND
(IAS15) including all eight planets, the Moon, and general-relativistic
corrections, validated against JPL Horizons ephemerides. Trained on
$\sim$2{,}000 real NEAs from the JPL Small-Body Database with MOID
$<0.05$~au, the surrogate reproduces the Monte-Carlo close-approach
distributions with well-calibrated uncertainties (CRPS, PIT, and coverage
diagnostics) at a $\sim$$10^{4-5}\times$ speed-up over direct $N$-body clone
propagation, enabling population-scale risk triage for surveys such as
Rubin/LSST, which are expected to discover $\sim$$10^5$ new NEAs.
\end{abstract}

\keywords{Near-Earth objects --- Asteroids --- Impact monitoring --- Neural networks}

\section{Introduction}

Assessing the impact hazard posed by a newly discovered near-Earth asteroid
(NEA) requires propagating the uncertainty of its orbit determination decades
into the future. Operational impact-monitoring systems such as JPL's Sentry
and Sentry-II \citep{Milani2005, Roa2021} sample the orbital-solution
covariance --- along the line of variations or in the full six-dimensional
element space --- and numerically integrate thousands of virtual asteroids
(``clones'') with high-fidelity dynamical models for $\sim$100~yr, recording
close approaches and impacts. The statistical distribution of the resulting
close-approach distances underpins the reported impact probabilities. This
procedure is computationally expensive: each newly discovered or newly
re-observed object requires thousands of long-term $N$-body integrations,
and the cost is incurred again after every orbit update.

The computational pressure is about to increase sharply. The Rubin
Observatory Legacy Survey of Space and Time (LSST) is expected to increase
NEA discovery rates by roughly an order of magnitude, ultimately cataloguing
on the order of $10^5$ NEAs \citep{Jones2018, Ivezic2019}. A fast,
reliable surrogate for uncertainty propagation would enable population-scale
risk triage, reserving full-fidelity Monte-Carlo (MC) analysis for the small
fraction of objects flagged as potentially hazardous.

Machine learning has so far touched this problem only peripherally.
\citet{Hefele2020} trained a multilayer perceptron to classify known
impactors versus non-impactors from tabulated orbital elements, without any
notion of orbit-determination uncertainty or predictive distributions;
subsequent works have largely revisited hazardous-object classification on
similar tabular data. To our knowledge, no published work learns the
\emph{uncertainty-propagation operator} itself: the map from an NEA's
orbital elements \emph{and} their covariance to the future probability
distribution of Earth close-approach distances.

Here we present such a surrogate. A mixture-density network
\citep[MDN;][]{Bishop1994} deep ensemble \citep{Lakshminarayanan2017} takes
as input the cometary orbital elements and $6\times6$ covariance of a real
NEA at its orbit-solution epoch and directly outputs the probability
distribution of the minimum Earth-approach distance in ten consecutive
decadal windows spanning 100~yr. Ground truth is generated by clone MC with
REBOUND \citep{ReinLiu2012} using the IAS15 integrator
\citep{ReinSpiegel2015} and general-relativistic corrections from REBOUNDx
\citep{Tamayo2020reboundx}. Section~\ref{sec:data} describes the data and
ground-truth generation, Section~\ref{sec:model} the surrogate,
Section~\ref{sec:results} the results, and Section~\ref{sec:discussion}
limitations and outlook.

\section{Data and Ground-Truth Generation}\label{sec:data}

\subsection{Target selection}

We query the JPL Small-Body Database (SBDB) for all 42{,}204 catalogued
near-Earth objects and retain those with orbit condition code $\leq 3$ and
observational data arc $\geq 365$~d (10{,}907 objects), ensuring
well-constrained orbital solutions. From these we select the 3{,}424 objects
with Earth MOID $< 0.05$~au --- the population for which close-approach
statistics are dynamically relevant --- and retrieve the full $6\times6$
covariance matrix of the cometary elements $(e, q, t_p, \Omega, \omega, i)$
at the orbit-solution epoch for each object via the SBDB per-object API.
The sample thus anchors the input distribution of the surrogate to the real
NEA population, both in the orbital elements and in the realistic structure
of orbit-determination covariances.

\subsection{$N$-body Monte-Carlo propagation}

Ground truth is generated with REBOUND using the adaptive high-order IAS15
integrator. The dynamical model comprises the Sun, the eight planets, and
the Moon, with initial conditions from JPL Horizons at epoch
JD~2461000.5, plus full general-relativistic corrections via the REBOUNDx
\texttt{gr} implementation. We validated the propagator against Horizons
ephemerides for (99942)~Apophis: the 10-yr position error is
$2.6\times10^{6}$~km without GR and $4.7\times10^{4}$~km with the full GR
term; the residual is attributable to the massive-asteroid perturbers and
the Yarkovsky acceleration included in the JPL dynamical model
\citep{Farnocchia2013} but not in ours.

For each target we draw clones from the multivariate normal distribution
defined by the SBDB covariance in $(e, q, t_p, \Omega, \omega, i)$ at the
covariance epoch and propagate them as massless test particles in a single
simulation per object (96 clones for training/validation objects, 256 for
test objects). Earth distances are sampled every 2.4~d (1.8~d for test
objects) and encounter minima are refined by quadratic interpolation. Each
clone is propagated for 100~yr, and we record the minimum Earth distance
$d_{\min}$ in each of ten consecutive 10-yr windows. Objects are assigned
to train/validation/test splits (80/8/12\%) by a deterministic hash of
their SPK identifier, so that no asteroid appears in more than one split.

\section{Neural Surrogate}\label{sec:model}

The input vector encodes the cometary elements ($e$, $q$, $\log a$;
angles as sine/cosine pairs; orbital phase at epoch as a sine/cosine pair),
three derived dynamical features ($\log_{10}$ Earth MOID, which is a hard
lower bound on $d_{\min}$; $\log_{10}$ orbital period; and the distance of
the period ratio to the nearest low-order mean-motion commensurability with
Earth), and the orbit-determination uncertainty as the signed-log-scaled
diagonal of the covariance Cholesky factor. The prediction window
$w \in \{0,\dots,9\}$ enters through a learned 16-dimensional embedding.

The network is a residual multilayer perceptron whose output head
parameterizes a $K$-component Gaussian mixture over $y = \log_{10}
(d_{\min}/\mathrm{au})$; mixture distributions accommodate the genuinely
multimodal outcomes produced by resonant-return dynamics. Training
minimizes the negative log-likelihood of individual MC clone outcomes
$(x, w) \mapsto y$, with early stopping on validation NLL. Epistemic
uncertainty is captured by a deep ensemble over independently initialized
training runs; the ensemble predictive distribution is the uniform mixture
of the member mixtures.

\section{Results}\label{sec:results}
% CRPS / NLL / W1 vs climatology baseline; calibration (PIT, coverage);
% quantile agreement; example distributions; speed-up benchmark;
% case studies (Apophis, Bennu); threshold reliability.
TODO

\section{Discussion and Conclusions}\label{sec:discussion}
TODO

\bibliography{refs}
\end{document}
